\begin{abstract}
Reinforcement Learning (RL), specifically Group Relative Policy Optimization (GRPO), has emerged as a transformative technique for unlocking deep reasoning in Large Language Models (LLMs). By incentivizing multi-step Chain-of-Thought (CoT) behaviors, GRPO allows models to transcend simple pattern matching. However, current implementations are plagued by critical bottlenecks: advantage vanishing in uniform groups, systematic correction bias during reflection, and inefficient utilization of noisy, high-entropy error samples. We propose \textbf{SB-GRPO} (Semantic-Balanced GRPO), an advanced framework that integrates dynamic multi-layer reflection with semantic error profiling. Our architecture introduces a differentiable \textbf{Heuristic Confidence Gate} that smoothly transitions between standard and negative-enhanced advantage calculation, preventing gradient stalls. We further employ K-Means clustering on reasoning embeddings to identify systematic logical failure modes and construct \textbf{Static Balanced Batches} to stabilize policy gradients. Finally, we derive an information-theoretic loss scaling factor based on the \textbf{Shannon Entropy} of semantic clusters to prioritize fixing systematic errors over random noise. Experimental evaluation on competitive benchmarks shows that SB-GRPO reaches a highly competitive accuracy of \textbf{74.80\%} on the MATH-500 dataset using a 7B model, outperforming vanilla GRPO by \textbf{26.40\%} and its multi-layer predecessors by \textbf{2.20\%}. Furthermore, it demonstrates substantial gains in elite logical reasoning, achieving up to \textbf{10.00\%} on AIME 2024.

\keywords{Reinforcement Learning \and Large Language Models \and Mathematical Reasoning \and Semantic Entropy}
\end{abstract}